\documentclass[fleqn,10pt]{wlscirep}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xcolor}

\usepackage{mathbbol}
\usepackage{multicol}
\usepackage[numbers,sort&compress]{natbib}

\newcommand{\NCMRGBR}{29,051\,}
\newcommand{\NCALLS}{801,526\,} % Number of genotyped SNPs 
\newcommand{\MAFTHR}{1\% } % MAF threshold (in percentage)
\newcommand{\NIMP}{9472708 } % number of % number of imputed variants with MAF > threshold
\newcommand{\HWEPVAL}{$10^{-5}$} % Hardy-Weinberg equilibrium p-value threshold
\newcommand{\LVNP}{2,677} % number of points in the LV mesh
\newcommand{\NGWASHITS}{10}

% ANONYMISED VERSION 
% \newcommand{\ACCESSIONNUMBER}{---}

\newcommand{\ACCESSIONNUMBER}{11350}

\title{Unsupervised deep learning helps disentangle the genetic basis of cardiac morphology}

\author[1,2,*]{Rodrigo Bonazzola}
\author[1,2]{Nishant Ravikumar}
\author[1,2]{Avan Suinesiaputra}
\author[6,7]{Bernard Keavney}
\author[2]{Sven Plein}
\author[3]{Enzo Ferrante}
\author[4]{Tanveer Syeda-Mahmood}
\author[1,2,5]{Alejandro F Frangi}
\affil[1]{Centre for Computational Imaging and Simulation Technologies in Biomedicine (CISTIB), School of Computing and School of Medicine, University of Leeds, Leeds, UK}
\affil[2]{Leeds Institute of Cardiovascular and Metabolic Medicine, School of Medicine, University of Leeds, Leeds, UK}
\affil[3]{Research Institute for Signals, Systems and Computational Intelligence, sinc(i), FICH-UNL / CONICET, Santa Fe, Argentina}
\affil[4]{IBM Almaden Research Center, San Jose, USA}
\affil[5]{Medical Imaging Research Center (MIRC), University Hospital Gasthuisberg. Cardiovascular Sciences and Electrical Engineering Departments, KU Leuven, Leuven, Belgium}
\affil[6]{Division of Cardiovascular Sciences, Faculty of Biology, Medicine and Health, University of Manchester, Manchester, UK}
\affil[7]{Manchester University NHS Foundation Trust, Manchester Academic Health Science Centre, Manchester, UK}

\affil[*]{scrb@leeds.ac.uk}


\begin{abstract}
Recent studies have identified associations between genetic variants and simple cardiac parameters like the volume of heart chambers or myocardial thickness. The emergence of big databases, including genetic data linked to cardiac magnetic resonance (CMR) images, with high spatial and temporal resolution facilitates investigation of more nuanced shape patterns. However, the high dimensionality of these images makes it challenging to find associations with genetic variants of potential clinical relevance. In this work, we exploit the UK Biobank database to extract left-ventricular geometric features from image-derived 3D meshes in an unsupervised fashion, via convolutional mesh autoencoders. These features are then tested in a genome-wide association study in order to unravel the genetic basis of cardiac morphology. Our results shed light on the role of gene phospholamban (PLN) in cardiac shape variation. In particular, our approach allowed to find a previously unreported link between this gene and left-ventricular sphericity.

% Recent studies have identified associations between genetic variants and simple cardiac parameters like the volume of heart chambers or myocardial thickness. The emergence of big databases, including genetic data linked to cardiac magnetic resonance (CMR) images, with high spatial and temporal resolution provides the means to describe more complex shape patterns. However, the high dimensionality of these images makes it challenging to find associations with genetic variants of potential clinical relevance. In this work, we propose a framework which combines geometric deep learning with genome-wide association studies to disclose novel links between cardiac anatomy and the underlying genetic basis. We construct convolutional mesh autoencoders to learn low-dimensional representations of cardiac meshes, and show that such embeddings encode anatomical variations linked to genetic differences. Our experimental results demonstrate that the proposed framework not only finds well-known genetic associations for cardiac shape changes, but also reveals novel links which open the door for future biomedical research.
\end{abstract}

\begin{document}

\flushbottom
\maketitle

\thispagestyle{empty}

% \input{legend}

\input{1_introduction}

\input{2_methods}

\input{3_results}

\input{4_conclusions}

% \section*{Discussion}

% Topical subheadings are allowed. Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work.

{\small
\bibliography{gwas_cardiomorpho.bib}
}

% \noindent LaTeX formats citations and references automatically using the bibliography records in your .bib file, which you can edit via the project menu. Use the cite command for an inline citation, e.g.  \cite{Hao:gidmaps:2014}.

% For data citations of datasets uploaded to e.g. \emph{figshare}, please use the \verb|howpublished| option in the bib entry to specify the platform and the link, as in the \verb|Hao:gidmaps:2014| example in the sample bibliography file.

\section*{Acknowledgements}

Acknowledgements should be brief, and should not include thanks to anonymous referees and editors, or effusive comments. Grant or contribution numbers may be acknowledged.

\section*{Author contributions statement}
R.B.,\newline 
N.R.,\newline 
A.S.,\newline 
T.S.M.,\newline 
E.F.,\newline 
A.F.,\newline 
All authors reviewed the manuscript. 

\section*{Additional information}

To include, in this order: \textbf{Accession codes} (where applicable); \textbf{Competing interests} (mandatory statement). 

The corresponding author is responsible for submitting a \href{http://www.nature.com/srep/policies/index.html#competing}{competing interests statement} on behalf of all authors of the paper. This statement must be included in the submitted article file.

% \appendix
\include{5_supplementary}
\end{document}


